Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-09-02 manu-draft.Rmd
more draft discussion

Make inventory of anomalous events..

Several of the correlations in Figure \@ref(fig:allcor) can be explained through mechanistic relationshps among water quality, macroalgae, and seagrasses.  Conversely, many of the correlations may be spurious and due to collinearity among variables that is likely not causative.  Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca). An increase in nitrogen was associated with an increase in chlorophyll, which in turn was associated with a decrease in water clarity.  Temporally, these changes were observed at different times and by different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa.  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen already utilized by different taxa, including macroalgae.

@@

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Although the seagrass community did not change dramatically across the study period, long-term impacts may be observed.  Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that peristed over a year and half.  These changes may have resulted from a reduction in water clarity from heavy concentrations of *K. brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.    Annually, seagrass coverage in Tampa Bay has declined... what other factors? PP is an affront...

>>>>

The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point together represent an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations provide a weight of evidence that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.  These anomalous events included 1) a large, spring diatom bloom in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for an unusual amount of observations supported by the decades of long-term monitoring data as a baseline for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.  In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.  Spill events reported to FDEP have been documented for years (e.g., industrial spills, stormwater pump failures, sanitary sewer overflows) and spill volumes for these for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point. Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.  From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in the Bay in 2021 is an outlier compared to past years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator and asset for environmental quality.

Several of the correlations in Figure \@ref(fig:allcor) support known responses of water quality to nutrient loads and can be explained through mechanistic relationships among water quality, macroalgae, and seagrasses.  Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca). An increase in nitrogen was associated with an increase in chlorophyll, which in turn was associated with a decrease in water clarity.  Temporally, these changes were observed at different times and by different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa.  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen already utilized by different taxa, including macroalgae.

@@

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.  Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year and half in southwest Florida.  These changes may have resulted from a reduction in water clarity from heavy concentrations of *K. brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.  Species shifts from macroalgal  to seagrass dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.  General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communites.  Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Managagement District.  From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest drivers acting at a regional scale are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.  Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
